### 2022-02-17 introduction/introduction.tex
jasmin

A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
would shape the future of automatic reasoning --- the resolution calculus
\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
elegant. It was also \emph{complete}: each theorem could be proven using this
system. Furthermore, it was less explosive than Davis and Putnam's algorithm
since it introduced unification as an inference filter.

@@

Superposition is a complete calculus for first-order logic loosely based on
resolution, featuring support for equality on the calculus level. This support
is further optimized by use of term orders and selection functions to prune the
search space. Thanks to efficient implementation, this calculus was used to
prove long-standing open problem known as ``Robbins conjecture''
\cite{mccune-97-robbins}. Provers based on this calculus have been
winning the first-order division of CASC theorem proving competition since its
inception in the 1990s \cite{ss-96-casc}.

@@

\paragraph{Proof Assistants and Hammers} Proof assistants (or interactive
theorem provers) are programs which allow users to describe a theory formally
and to write proofs of the statements about this theory using inference 
rules of the proof assistant. These programs can then check if the given proof
is correct. As their core is usually small and well-tested, proofs that pass the
checking phase are considered trustworthy.

Proof assistants showed indispensible in proving very complex mathematical theorems in a
trustworthy manner. For example, four-color theorem \cite{gg-2008-four-color}
and Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
ending the need to trust hundreds of pages of hard mathematical proofs. Proof
assistants are also used to show correctness of complex software such as
compilers \cite{xl-09-compcert} or operating systems kernels \cite{gk-09-sel4}.
This verified software can be used in safety-critical scenarios such as nuclear
plants or autonomous vehicles.

@@

Integrating proof automation right into the proof assistant is not an easy task.
Most proof assistants are based on variants of higher-order logic, which is more
expressive than first-order logic, but harder to automate as in general it does not even allow complete
proof calculi. To provide some level of automation, proof assistant developers
use the efficient first-order provers as follows. First, the proof goal is
translated from higher-order to first-order logic. Then the first-order prover
is run on the translated problem. If the first-order prover finds the proof, the
proof is reconstructed in the proof assistant. This approach is implemented in
modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
and Isabelle \cite{lc-88-isabelle} using \emph{hammers} CoqHammer
\cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
\cite{bn-10-sh}, respectively.

The approach proved useful: around 76\% of the simple proof goals can be
discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. Thus, from a standpoint of proof assistant ideal
automatic theorem prover fulfills the following wishlist:
\begin{enumerate}
  \item\label{it:fo} Performs as best first-order provers on first-order problems
  \item\label{it:size} Scales well with the size of the problem
  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
\end{enumerate}
This brings us to the topic of this thesis: \emph{How to develop a prover fulfilling this wishlist?}

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength --- from state-of-the-art
first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
prover to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist are already fulfilled, as E is very
efficient and deals well with large problems.

@@

We have split the road to full higher-order logic in three stages. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
disprove this statement. The first stage is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. Suppose we want to prove that there is a
function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
function can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stage is to support automatically synthesizing such functions.

In higher-order logic we can make statements about statements. More precisely,
it treats Boolean terms (formulas) as first-class citizens, enabling us to
model some concepts in more natural manner. For example, we can naturally
represent statements such as ``what Jasmin says is true'' as $\forall x. \,
\cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support Boolean terms
as first-class citizens.
\pagebreak[2]

@@

The biggest contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend those provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term representation of Ehoh to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and less redundant than current state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, strong basis that we have earlier built made it
  manageable.

  \item Following our approach of lifting efficient provers from
  first-order to higher-order logic, we went a step back in an effort to improve
  first-order provers. In particular, we looked at the most efficient approaches
  to simplify problems in propositional logic, a simpler, decidable logic that disallows quantification. 
  We lifted some of these approaches to first-order logic
  and implemented them in Zipperposition. 
\end{itemize}

@@

This thesis is cumulative: chapters~\ref{ch:ehoh}~-~\ref{ch:satfol}  are
directly taken from previous publications (or drafts in case of
Chapter~\ref{ch:ehoh2}) where I was the main author. Use
of publications in this thesis was authorized by the coauthors.

Chapter~\ref{ch:pre} introduces the background material necessary for following this thesis,
while Chapter~\ref{ch:conclusion} gives a brief summary of the work and describes alleys for future work.
Other six chapters discuss previously described contributions in detail:

\vspace{0.5em}
\noindent\textbf{Chapter~\ref{ch:ehoh}} discusses the steps we have taken to extend first-order
  prover E to $\lambda$-free higher-order logic, obtaining Ehoh. It is mostly based on the journal publication
  \begin{enumerate}
    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
    In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021.  
  \end{enumerate}
\vspace{-0.3em}
while parts of the chapter were previously published in the conference publication
\vspace{-0.3em}
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
    In  In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
  \end{enumerate}
\noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification algorithm we designed to be used
inside efficient theorem provers. It is mostly based on the journal publication
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
    In \emph{Logical Methods in Computer Science} 17(4): 18:1--18:31, 2021.
  \end{enumerate}
\vspace{-0.3em}
while parts of the chapter were previously published in the conference
publication, which won the best paper by junior researcher award:
\vspace{-0.3em}
\begin{enumerate}[resume]
  \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
  In Ariola, Z.M. (ed.) \emph{FSCD 2020} LIPIcs 167, pp. 5:1--5:17, Schloss Dagstuhl--Leibniz-Zentrum für Informatik, 2020. 
\end{enumerate}
\noindent\textbf{Chapter~\ref{ch:bools}} discusses the pragmatic extensions of
the complete calculus for superposition with lambdas to support first-class
Booleans. It is based on the publication
    \begin{enumerate}[resume]
      \item Petar Vukmirović and Visa Nummelin. Boolean Reasoning in a Higher-Order
      Superposition Prover. In P. Fontaine, K. Korovin, I.S. Kotsireas, P.
      R\"ummer, S. Tourret (eds.) \emph{PAAR-2020}, CEUR Workshop Proceedings, vol.
      2752, pp. 148--166. CEUR-WS.org (2020)
    \end{enumerate}
\noindent\textbf{Chapter~\ref{ch:ho-techniques}} discusses the approach we took to
control explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
    Making Higher-Order Superposition Work. In \emph{Journal of Automated Reasoning}. Springer, 2022.
  \end{enumerate}
\vspace{-0.3em}
  while parts of the chapter were previously published in the conference publication, which won the best paper by a student award:
\vspace{-0.3em}
\begin{enumerate}[resume]
  \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
  Making Higher-Order Superposition Work. In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28},
  LNCS 12699, pp. 415--432, Springer, 2021.
\end{enumerate}
\noindent\textbf{Chapter~\ref{ch:ehoh2}} discusses the extension of Ehoh to
support full higher-order logic. We call the obtained prover \newehoh.
It is based on the draft 
\begin{enumerate}[resume]
  \item Petar Vukmirović, Jasmin Blanchette, Stephan Schulz.  Extending a
  High-Performance Prover to Higher-Order Logic. 2022.
\end{enumerate}

@@

These attempts resulted in early higher-order theorem provers. Andrews and
collegaues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
and colleageues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced cooperative paradigm. Provers of this paradigm
invoke first-order back-ends in regular intervals to finish the proof. Latest
iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
to close tableau branches. It won higher-order division of CASC on 7 ocassions:
in 2011 and every year from 2013 to 2019.

Various reserachers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

Higher-order unification is one of the central procedures in a higher-order
prover. Jensen and Pietrzykowski introduced a complete procedure for enumerating
elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
some calculi remaing complete if solving a difficult subproblem of higher-order
unification (``flex-flex'' pairs) is delayed. He described procedure to solve
this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
desgined a procedure to enumerate higher-order unifiers using higher-order
combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat
and Reger implemented efficient term indexing in their extension of Vampire
\cite{br-20-full-sup-w-combs}. Libal and Steen designed efficient term indexing
based on substitution trees \cite{ls-16-indexing}.

There have been many studies on the performance of algorithms or success rate of
heuristics in first-order theorem proving. For example, Hoder and Voronkov
evaluate different unification algorithms \cite{hv-09-unifalgs}, while Reger and
Voronkov evaluate different parameters of AVATAR architecture
\cite{rsv-15-playing-with-avatar}. Other authors evaluated different proof
search heuristics for superposition provers \cite{gs-20-clausesel,
hrsv-16-selsel, sm-16-clausesel}. Evaluation of many prover parameters has been
done for higher-order provers as well
\cite{benzmueller-et-al-05-can-ho-fo-coop,fb-2016-internal-guidance-satallax,sb-15-beta,wskb-16-effective-norm}.

>>>>

A bigger breakthrough happened in 1965 when Robinson introduced a calculus that
would shape the future of automatic reasoning --- the resolution calculus
\cite{ar-65-resolution}. Consisting of a single inference rule it was simple and
elegant. It was also \emph{complete}: each theorem could be proven using this
system. Furthermore, it was less explosive than Davis and Putnam's algorithm
since it provided better guided search space exploration.

@@

Superposition is a complete calculus for first-order logic loosely based on
resolution, featuring support for equality on the calculus level. This support
is further optimized by use of various heuristics built to help prune the
search space. Thanks to efficient implementation, this calculus was used to
prove long-standing open problem known as the Robbins conjecture
\cite{mccune-97-robbins}. Provers based on this calculus have been
winning the first-order division of the CASC theorem proving competition since its
inception in the 1990s \cite{ss-96-casc}.

@@

\paragraph{Proof Assistants and Hammers} Proof assistants (also called interactive
theorem provers) are programs which allow users to describe a theory formally
and to write proofs of the statements about this theory using inference 
rules of the proof assistant. These programs can then check if the given proof
is correct. As their core is usually small and well-tested, proofs that pass the
checking phase are considered trustworthy.

Proof assistants were essential in proving very complex mathematical theorems in a
trustworthy manner. For example, the four-color theorem \cite{gg-2008-four-color}
and the Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
ending the need to trust hundreds of pages of mathematical proofs. Proof
assistants are also used to show correctness of complex software such as
compilers \cite{xl-09-compcert} or operating systems kernels \cite{gk-09-sel4}.
This verified software can be used in safety-critical scenarios such as nuclear
plants or autonomous vehicles.

@@

Integrating proof automation directly into the proof assistant is not an easy task.
Most proof assistants are based on variants of higher-order logic, which is more
expressive than first-order logic, but harder to automate as in general it does not even allow complete
proof calculi. To provide some level of automation, proof assistant developers
use the efficient first-order provers as follows. First, the proof goal is
translated from higher-order to first-order logic. Then the first-order prover
is run on the translated problem. If the first-order prover finds the proof, the
proof is reconstructed in the proof assistant. This approach is implemented in
modern assistants such as Coq \cite{bc-04-coq}, HOL Light \cite{jh-09-hol-light}
and Isabelle \cite{lc-88-isabelle} using the \emph{hammers} CoqHammer
\cite{ck-18-coqhammer}, HOL(y)Hammer \cite{ku-15-holyhammer} and Sledgehammer
\cite{bn-10-sh}, respectively.

The approach proved useful: on some benchmarks, around 77\% of the proof goals can be
discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. Thus, from the standpoint of proof assistants the ideal
automatic theorem prover fulfills the following wishlist:
\begin{enumerate}
  \item\label{it:fo} Performs as the best first-order provers on first-order problems
  \item\label{it:size} Scales well with the size of the problem
  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
\end{enumerate}
This brings us to the topic of this thesis: \emph{How to develop a prover fulfilling this wishlist?}

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic, we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength --- from the state-of-the-art
first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
prover to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist are already fulfilled, as E is very
efficient and deals well with large problems.

@@

We identified three stops on the road to full higher-order logic. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. \,  x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with the addition function ($\cst{+}$) is enough to
disprove this statement. The first stop is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. Suppose we want to prove that there is a
function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
function can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stop is to support automatically synthesizing such functions.

In higher-order logic we can make statements about statements. More precisely,
it treats Boolean terms (formulas) as first-class citizens, enabling us to
model some concepts in more natural manner. For example, we can naturally
represent statements such as ``what Jasmin says is true'' as $\forall x. \,
\cst{says}(\cst{jasmin}, x) \imp x$. The last stop is to support Boolean terms
as first-class citizens.
\pagebreak[2]

@@

The main contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend these provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve the best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term representation of E to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It finished third, 1 percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and generated less redundant unifiers than the state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 won the higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points ahead of the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, the strong basis that we have earlier built made it
  manageable.

  \item After spending so much time and energy on developing higher-order
  provers, we took a step back and worked on improving first-order provers. In
  particular, we looked at the most efficient approaches to simplify problems in
  propositional logic, a simpler, decidable logic that disallows quantification.
  We lifted some of these approaches to first-order logic and implemented them
  in Zipperposition. 
\end{itemize}

@@

This thesis is cumulative: chapters~\ref{ch:ehoh}--\ref{ch:satfol}  are
taken from previous publications (or drafts in case of
Chapter~\ref{ch:ehoh2}) where I was the main author. Use
of the publications in this thesis was authorized by the coauthors.

Chapter~\ref{ch:pre} introduces the background material necessary for following this thesis,
while Chapter~\ref{ch:conclusion} gives a brief summary of the work and describes alleys for future work.
The other six chapters discuss previously described contributions in detail:

\vspace{0.5em}
\noindent\textbf{Chapter~\ref{ch:ehoh}} discusses the steps we have taken to extend first-order
  prover E to $\lambda$-free higher-order logic, obtaining Ehoh. It is mostly based on the journal publication
  \begin{enumerate}
    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
    In \emph{International Journal on Software Tools for Technology Transfer}. Springer, 2021. Published online only.
  \end{enumerate}
\vspace{-0.3em}
while parts of the chapter were previously published in the conference publication
\vspace{-0.3em}
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Jasmin Blanchette, Simon Cruanes, and Stephan Schulz.
    Extending a Brainiac Prover to Lambda-Free Higher-Order Logic.
    In Vojnar, T., Zhang, L. (eds.) \emph{TACAS 2019}, LNCS 11427, pp. 192--210, Springer, 2019.
  \end{enumerate}
\noindent\textbf{Chapter~\ref{ch:unif}} discusses the higher-order unification procedure we designed to be used
inside efficient theorem provers. It is mostly based on the journal publication
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
    In \emph{Logical Methods in Computer Science} 17(4): 18:1--18:31, 2021.
  \end{enumerate}
\vspace{-0.3em}
while parts of the chapter were previously published in the conference
publication, which won the best paper by a junior researcher award:
\vspace{-0.3em}
\begin{enumerate}[resume]
  \item Petar Vukmirović, Alexander Bentkamp, and Visa Nummelin. Efficient Full Hi\-gher-Order Unification. 
  In Ariola, Z.M. (ed.) \emph{FSCD 2020} LIPIcs 167, pp. 5:1--5:17, Schloss Dagstuhl--Leibniz-Zentrum für Informatik, 2020. 
\end{enumerate}
\noindent\textbf{Chapter~\ref{ch:bools}} discusses the pragmatic extensions of
the complete calculus for superposition with $\lambda$-abstractions to support first-class
Booleans. It is based on the publication
    \begin{enumerate}[resume]
      \item Petar Vukmirović and Visa Nummelin. Boolean Reasoning in a Higher-Order
      Superposition Prover. In P. Fontaine, K. Korovin, I.S. Kotsireas, P.
      R\"ummer, S. Tourret (eds.) \emph{PAAR-2020}, CEUR Workshop Proceedings, vol.
      2752, pp. 148--166. CEUR-WS.org (2020)
    \end{enumerate}
\noindent\textbf{Chapter~\ref{ch:ho-techniques}} discusses the approach we took to
control the explosion of the search space inherent to higher-order superposition. It is mostly based on the journal publication
  \begin{enumerate}[resume]
    \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
    Making Higher-Order Superposition Work. In \emph{Journal of Automated Reasoning}. Springer, 2022.
  \end{enumerate}
\vspace{-0.3em}
  while parts of the chapter were previously published in the conference publication, which won the best paper by a student award:
\vspace{-0.3em}
\begin{enumerate}[resume]
  \item Petar Vukmirović, Alexander Bentkamp, Jasmin Blanchette, Simon Cruanes, Visa Nummelin, and Sophie Tourret.
  Making Higher-Order Superposition Work. In Platzer, A., Sutcliffe, G. (eds.) \emph{CADE-28},
  LNCS 12699, pp. 415--432, Springer, 2021.
\end{enumerate}
\noindent\textbf{Chapter~\ref{ch:ehoh2}} discusses the extension of Ehoh to
support full higher-order logic, resulting in \newehoh.
It is based on the draft 
\begin{enumerate}[resume]
  \item Petar Vukmirović, Jasmin Blanchette, and Stephan Schulz.  Extending a
  High-Performance Prover to Higher-Order Logic. 2022.
\end{enumerate}

@@

These attempts resulted in early higher-order theorem provers. Andrews and
colleagues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
and colleagues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced the cooperative paradigm. Provers implementing this paradigm
invoke first-order backends in regular intervals to finish the proof. Latest
iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
to close tableau branches. It won higher-order division of CASC on eight occasions:
in 2011 and every year from 2013 to 2019.

Various researchers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

Higher-order unification is one of the central procedures in a higher-order
prover. Jensen and Pietrzykowski introduced a complete procedure for enumerating
elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
some calculi remaing complete if solving a difficult subproblem of higher-order
unification (``flex-flex'' pairs) is delayed. He described procedure to solve
this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
desgined a procedure to enumerate higher-order unifiers using higher-order
combinators~\cite{dd-93-comb-unif}.

To make unification more efficient, Bhayat
and Reger implemented efficient term indexing in their extension of Vampire
\cite{br-20-full-sup-w-combs}. Libal and Steen designed efficient term indexing
based on substitution trees \cite{ls-16-indexing}.

There have been many studies on the performance of algorithms or success rate of
heuristics in first-order theorem proving. For example, Hoder and Voronkov
evaluate different unification algorithms \cite{hv-09-unifalgs}, while Reger and
Voronkov evaluate different parameters of a novel prover architecture
\cite{rsv-15-playing-with-avatar}. Other authors evaluated different proof
search heuristics for superposition provers \cite{gs-20-clausesel,
hrsv-16-selsel, sm-16-clausesel}. Evaluation of many prover parameters has been
done for higher-order provers as well
\cite{benzmueller-et-al-05-can-ho-fo-coop,fb-2016-internal-guidance-satallax,sb-15-beta,wskb-16-effective-norm}.
